Variable Density Tesseroids: Gravity fields calculation in spherical coordinates using variable densities in depth

### 2017-10-19 manuscript/manuscript.tex
Changes in manuscript

We have enhanced the later implementation with a new forward gravity model that allows us to compute the gravity fields generated by any tesseroid with an arbitrary density function on any external point.
It is based on the GLQ approximation, makes use of the modified distance-size discretization algorithm and a new density-based discretization that increases the accuracy for high variating density functions.
It is written in Python and Cython languages, constituting an easy to use library that runs precompiled code for the more time consuming functions.

@@

In the following sections we will describe how the new algorithm works, its theoretical background and obtain the distance-size ratio values needed to get a good accuracy in the computation.

@@

\begin{equation}
    \int\limits_{-1}^1 f(x) dx \approx \frac{b-a}{2} \sum_{i=1}^N W_i f(x_i),
\end{equation}

@@

From equation \ref{eq:glq-var-dens} we can observe that the GLQ approximates the gravity fields of a tesseroid with variable density as the ones generated by point masses located at the nodes of the Legendre polynomials.
Furthermore, the information of the density function is summarised as the values it assumes on those nodes.
Although this may sound discouraging, we intent to prove otherwise and show that this method, along with a well fitted discretization algorithm, approximates the gravity fields with good accuracy and precision.\todo{no se si poner esta ultima oracion}

@@

\noindent and the integration in $\phi'$ can be performed independently of the density function.
Using \citet{sagemath} we obtain the following expression:

@@

The equation \ref{eq:shell-pot} allows to easily obtain the exact gravity field generated by a spherical shell with a variable density in depth on any outside point located in the $z$ axe.
Due to the rotational symmetry along any axe that passes through the centre of the shell, the integral in equation \ref{eq:shell-pot} reproduces the gravity potential field on any outside point at distance $r$ from the its centre.

The gradient and the Marussi tensor derived from potentials that depends solely on $r$ has only a few non zero components: the vertical component of the gradient ($g_z$) and the diagonal components of the tensor ($g_{xx}$, $g_{yy}$, $g_{zz}$).
They can be easily obtained as \citet{Grombein2013} does, even for any density function $\rho(r')$ of our choice:

@@

Due to a new type of approximation error resulting from high variations in the density functions we had to develop a new density-based discretization algorithm that also intent to increase the accuracy, working together with the preexisting adaptive discretization algorithm.

@@

On the other hand, the value assigned to the distance-size ratio $D$ cannot be easily related to the error of the approximation, thus the choice of the value of $D$ to get an acceptable accuracy is a matter of study.

In order to overcome this, \citet{Uieda2016} compared the numerical computation with the analytical solution for a spherical shell with homogeneous density.
We have performed similar tests, but now with variable density spherical shell.
From the infinite density functions that can be proposed, we have chosen three characteristic ones: a linear, an exponential and a discontinuous function.

@@

High variations in the density function of an arbitrary tesseroid introduces a new type of error into the numerical approximation: only a few point masses would not reproduce the effect of this kind of variations.
The modified version of the adaptive discretization algorithm may help in reducing this kind of error, but as it takes into account only the dimensions of the tesseroid and the distance to the computation point, it's not well suited to fully perform this task.

@@

\begin{enumerate}
\renewcommand{\theenumi}{(\arabic{enumi})}
    \item Normalise the density function to the [0, 1] interval,
    \item Compute the absolute difference between the normalised density and a linear function that assumes the same values on the top and bottom radii that this normalised density.
    \item If the maximum of the absolute difference is greater that a predefined delta ratio ($\delta$), then the tesseroid will be subdivided on the radius at which this maximum takes place. Otherwise, the tesseroid won't be subdivided.
\end{enumerate}

First we need to normalise the density function such as it assumes the values of 0 and 1 on the boundary radii ($R_1$, $R_2$), respectively, depending at which radius the original density function is maximum.
We call this normalised density $\rho_n(r')$ and we can calculate it as follows:

@@

This algorithm can be subsequently applied to every subdividing tesseroid, until the inequation \ref{eq:delta-density} is not held or the subdivisions are smaller than 1mm.
This two conditions guarantee that the algorithm is finite and does not fall into an infinite loop. \todo{chequear esta oracion}
Once this conditions are met, every subdividing tesseroid is subjected to the modified adaptive discretization algorithm and the GLQ in order to compute the gravity field generated by it.

It's easy to see that the delta ratio controls how many times the tesseroids will be subdivided based on the variation of the density function and therefore it controls the accuracy involving the numerical error that the density variation introduces.
This raise the need of determine a maximum value of $\delta$ that ensures an acceptable accuracy.
We will perform this determination with the same methodology as the one applied to the $D$ ratio: compare the numerical model with the analytical solutions for a spherical shell.

@@

This steps are applied for each computation point, adding the contribution of each tesseroid to the resulting gravity field.

@@

In order to determine the minimum value for the distance-size ratio $D$ such the numerical approximation achieves an acceptable accuracy we have compared the numerical model with the analytical solutions for a spherical shell with variable density for three special cases: a linear, an exponential and a discontinuous density function.
The first one constitutes the simplest variation, while the second one represents the fastest varying function. The later one tries to emulate a contact between two different media (e.g. crust and mantle interface).

We approximated the spherical shell by a mesh of $30^\circ \times 30^\circ$ tesseroids and calculated the gravity potential, the vertical component of the gradient ($g_z$) and the diagonal components of the Marussi tensor ($g_{xx}$, $g_{yy}$, $g_{zz}$) on four different grids: (1) a grid located at the pole, (2) another one on the Equator, (3) one located at satellite height and (4) a big grid of $30^\circ \times 30^\circ$.
More information about this grids can be found in the Table \ref{tab:grids}.

@@

\begin{equation}
    \rho(r') = A \left( e^{-(r' - R)/b} - 1\right) + 2670 \text{kg/m}^3
\label{eq:density-exp}
\end{equation}
\noindent where
\begin{equation}
    A =
    (3300 \text{kg/m}^3 - 2670 \text{kg/m}^3)
    \left( e^{( R_2 - R_1 )/b} - 1 \right)^{-1},
\end{equation}

\noindent and $b$ = 1000km. 
In this way, the shell has a density of 2670 kg/m$^3$ and 3300 kg/m$^3$ on its outer and inner radii, respectively.
The analytical solution of the gravity potential is easily obtained as the sum the solution for a pure exponential function and the one for an homogeneous spherical shell.

We have performed this test also with a thin and a thick shell of 1 km and 35 km, respectively, without applying the density-based discretization algorithm. Its results can be seen in Figures \ref{fig:D-exp-thin} and \ref{fig:D-exp-thick}.
Once again, the results for $g_{xx}$ and $g_{yy}$ are very similar to the ones obtained for $g_{zz}$, so we exclude them, although, they can be found in the repository.

In the same way we analysed the linear density results, we can summarise that the thin spherical shell presents the more conservative values of the distance-size ratio: $D=1$for the gravity potential, $D=2$ for its gradient components and $D=8$ for the Marussi tensor components.

Surprisingly both the linear and the exponential density tesseroid models needs the same values of distance-size ratio $D$ in order to assure a good accuracy.

@@

Because the constant $b$ controls the variation of the density function, we wanted to test how does the accuracy of the method behaves at different values of $b$ when the density-based discretization algorithm is not applied.
In order to do this we proposed a set of density functions, each one with a different value of $b$.
We computed the difference between the Marussi tensor component $g_{zz}$ generated by the numerical model on the big grid of 30$^\circ\times$30$^\circ$ (see Table \ref{tab:grids}) and the respective analytical solution.
These differences were calculated for each proposed density function and for the same set of values of $D$ we have been using on the previous tests, obtaining one curve for each value of $b$.

On Figure \ref{fig:D-exp-power-thin}a we can see the density functions used to perform these tests and its corresponding results on \ref{fig:D-exp-power-thin}b .
Taking into account that lower values of $b$ implies a more varying function, we can see that for high varying functions the numerical model does not perform well: for $b$=100m the difference is almost 100\%.

@@

\begin{equation}
    \rho(r') = 
    \begin{cases}
        \rho_1 & \text{if} \, R_1 \leq r' < R_c \\
        \rho_2 & \text{if} \, R_c \leq r' \leq R_2 \\
    \end{cases}
\label{eq:density-discontinuous}
\end{equation}

\noindent where $\rho_1$ and $\rho_2$ are constant densities, $R_1$ and $R_2$ are the inner and outer radii of the shell, respectively, and $R_c$ is the the radius on which the density change takes place.

The analytical solution for this kind of density is a linear combination of two homogeneous shells with densities $\rho_1$ and $\rho_2$:

\begin{equation}
    V(r) = \frac{4}{3} \frac{\pi G}{r}
    \left[ \rho_1 (R_c^3 - R_1^3) +
    \rho_2 (R_2^3 - R_c^3) \right]
\end{equation}

@@

>>>>

In addition, the varying density inside an arbitrary tesseroid introduces a new kind of error into the numerical approximation.
This encouraged us to develop a new density-based discretization algorithm that divides the tesseroid into smaller ones taking into account how the density function varies.
The number of subdivisions and thus the accuracy of the approximation will be controlled by a delta ratio $\delta$, a predefined value by the user.

We have enhanced the implementation of \citet{Uieda2016} with a new forward gravity model that allows us to compute the gravity fields generated by any tesseroid with an arbitrary density function on any external point.
It is based on the GLQ approximation, makes use of the modified distance-size discretization and the new density-based discretization algorithms that guarantee the accuracy even for high varying density functions.
It is written in Python and Cython languages, constituting an easy to use library that runs precompiled code for the more time consuming functions.

@@

In the following sections we will describe how the new algorithm works, its theoretical background and obtain default distance-size ratio and delta ratio values needed to get an acceptable accuracy in the numerical approximation.

@@

\begin{equation}
    \int\limits_{-1}^1 f(x) dx \approx \sum_{i=1}^N W_i f(x_i),
\end{equation}

@@

From equation \ref{eq:glq-var-dens} we can observe that the GLQ approximates the gravity fields of a tesseroid with variable density as the ones generated by point masses located at the nodes of the Legendre polynomials.
Furthermore, we can also conclude that the information of the density function is summarised as the values it assumes on those nodes.
Although this may sound discouraging, we intent to prove otherwise and show that this method, along with well fitted discretization algorithms, approximates the gravity fields with good accuracy.\todo{no se si poner esta ultima oracion}

@@

\noindent while the integration in $\phi'$ can be performed independently of the density function.
Using \citet{sagemath} we obtained the following expression:

@@

The equation \ref{eq:shell-pot} allows to easily obtain the exact gravity potential generated by a spherical shell with a variable density in depth on any outside point located in the $z$ axe.
Due to the rotational symmetry along any axe that passes through the centre of the shell, the integral in equation \ref{eq:shell-pot} reproduces the gravity potential field on any outside point at distance $r$ from the its centre.

The gradient and the Marussi tensor derived from potentials that depends solely on $r$ have only a few non zero components: the vertical component of the gradient ($g_z$) and the diagonal components of the tensor ($g_{xx}$, $g_{yy}$, $g_{zz}$).
They can be easily obtained as \citet{Grombein2013} does, even for any arbitrary density function $\rho(r')$:

@@

In addition, we had to develop a new density-based discretization algorithm in order to decrease a new type of error in the numerical approximation caused by high variations in the density functions.
It is intended to work together with the preexisting modified adaptive discretization algorithm.

@@

On the other hand, the value assigned to the distance-size ratio $D$ cannot be easily related to the error of the approximation, thus the choice of the value of $D$ to get an acceptable accuracy must be objectively determined beforehand.
In order to overcome this, \citet{Uieda2016} compared the numerical computation with the analytical solution for a spherical shell with homogeneous density.

We will perform similar tests, but now with variable density spherical shell.
From the infinite density functions that can be proposed, we have chosen three characteristic ones: a linear, an exponential and a discontinuous function.

@@

High variations in the density function of an arbitrary tesseroid introduces a new type of error into the numerical approximation: only a few point masses would not reproduce the effect of this kind of variations.
The modified version of the adaptive discretization algorithm may help in reducing this kind of error, but as it only takes into account the dimensions of the tesseroid and the distance to the computation point, it's not well suited to fully perform this task.

@@

\begin{enumerate}
\renewcommand{\theenumi}{(\arabic{enumi})}
    \item Normalise the density function to the [0, 1] interval,
    \item Compute the absolute difference between the normalised density and a linear function that assumes the same values on the top and bottom radii that this normalised density.
    \item If the maximum of the absolute difference is greater that a predefined delta ratio ($\delta$), then the tesseroid will be subdivided on the radius at which this maximum takes place. Otherwise, the tesseroid won't be split.
\end{enumerate}

First we need to normalise the density function such as it assumes the values of 0 and 1 on the boundary radii ($R_1$, $R_2$), depending at which radius the original density function is maximum.
We call this normalised density $\rho_n(r')$ and we calculate it as follows:

@@

This algorithm can be subsequently applied to every subdividing tesseroid, until the inequation \ref{eq:delta-density} is not held or the subdivisions are smaller than 1mm.
This two conditions guarantee that the algorithm is finite, i.e. it does not fall into an infinite loop. \todo{chequear esta oracion}
Once this conditions are met, every subdividing tesseroid is subjected to the modified adaptive discretization algorithm and the GLQ in order to compute the gravity field generated by it.

It's easy to see that the delta ratio indirectly controls how many times the tesseroids will be subdivided based on the variation of the density function. Therefore it controls the accuracy involving the numerical error that the density variation introduces and also the computation time.
This raises the need to determine a maximum value of $\delta$ that ensures an acceptable accuracy while minimising the computation time.
We will perform this determination with the same methodology as the one applied to the $D$ ratio: compare the numerical model with the analytical solutions for a spherical shell.

@@

These steps are applied for each computation point, adding the contribution of each tesseroid to the resulting gravity field.

@@

In order to determine the minimum value for the distance-size ratio $D$ such the numerical approximation achieves an acceptable accuracy we have compared the numerical model with the analytical solutions for a spherical shell with variable density for three special cases: a linear, an exponential and a discontinuous density function.
The first one constitutes the simplest variation, while the second one represents a higher varying function. The later one tries to emulate a contact between two different media (e.g. crust and mantle interface).

We approximated the spherical shell by a mesh of $30^\circ \times 30^\circ$ tesseroids and calculated the gravity potential, the vertical component of the gradient ($g_z$) and the diagonal components of the Marussi tensor ($g_{xx}$, $g_{yy}$, $g_{zz}$) on four different grids: (1) a grid located at the pole, (2) another one on the Equator, (3) one located at satellite height and (4) a big grid of $30^\circ \times 30^\circ$.
More information about this grids can be found on Table \ref{tab:grids}.

@@

\begin{equation}
    \rho(r') = A \left( e^{-(r' - R)/b} - 1\right) + 2670 \text{kg/m}^3,
\label{eq:density-exp}
\end{equation}
\noindent where
\begin{equation}
    A =
    (3300 \text{kg/m}^3 - 2670 \text{kg/m}^3)
    \left( e^{( R_2 - R_1 )/b} - 1 \right)^{-1},
\end{equation}

\noindent and the value of $b$ depends on the spherical shell on which we performed the test. 
In this way, the shell has a density of 2670 kg/m$^3$ and 3300 kg/m$^3$ on its outer and inner radii, respectively.
The analytical solution of the gravity potential is easily obtained as the sum the solution for a pure exponential function and the one for an homogeneous spherical shell.

We have performed this test also with a thin and a thick shell of 1 km and 35 km, respectively, without applying the density-based discretization algorithm.
On both cases, the constant $b$ is equal to the thickness of the shell: 1km for the thin shell and 35km for the thick one.
The results can be seen in Figures \ref{fig:D-exp-thin} and \ref{fig:D-exp-thick}.
Once again, the results for $g_{xx}$ and $g_{yy}$ are very similar to the ones obtained for $g_{zz}$, so we exclude them, although, they can be found in the repository.

In the same way we analysed the linear density results, we can summarise that the thin spherical shell presents the more conservative values of the distance-size ratio: $D=1$ for the gravity potential, $D=2$ for its gradient components and $D=8$ for the Marussi tensor components.

@@

Because the constant $b$ controls the variation of the density function, we wanted to test how does the accuracy of the method behaves at different values of $b$ when the density-based discretization algorithm is not applied.
In order to do this we proposed a set of density functions, each one with a different value of $b$.
We computed the difference between the Marussi tensor component $g_{zz}$ generated by the numerical model on the big grid of 30$^\circ\times$30$^\circ$ (see Table \ref{tab:grids}) and the corresponding analytical solution.
These differences were calculated for each proposed density function and for the same set of values of $D$ we have been using on the previous tests, obtaining one curve for each value of $b$.

On Figure \ref{fig:D-exp-power-thin}a we can see the density functions used to perform these tests and its corresponding results on \ref{fig:D-exp-power-thin}b .
Taking into account that lower values of $b$ implies a more varying function, we can see that for high varying functions the numerical model does not perform well: for $b=100$m the difference is almost 100\%.

@@

\begin{equation}
    \rho(r') = 
    \begin{cases}
        \rho_A & \text{if} \, R_1 \leq r' < R_c \\
        \rho_B & \text{if} \, R_c \leq r' \leq R_2 \\
    \end{cases}
\label{eq:density-discontinuous}
\end{equation}

\noindent where $\rho_A$ and $\rho_B$ are constant densities, $R_1$ and $R_2$ are the inner and outer radii of the shell, respectively, and $R_c$ is the the radius on which the density change takes place.

The analytical solution for this kind of density is a linear combination of two homogeneous shells with densities $\rho_A$ and $\rho_B$:

\begin{equation}
    V(r) = \frac{4}{3} \frac{\pi G}{r}
    \left[ \rho_A (R_c^3 - R_1^3) +
    \rho_B (R_2^3 - R_c^3) \right]
\end{equation}

@@

\section{Determination of delta ratio}

In the light of the previous results, a density-based discretization algorithm is needed in order to achieve an acceptable accuracy in those cases on which the modified adaptive discretization algorithm is not sufficient to take into account the numerical error due to density variations.

Besides the implementation of this new algorithm, a determination of the maximum value of the delta ratio $\delta$ that produces an accuracy bellow a predefined threshold is also needed in order to guarantee the good performance of the numerical model.

To do this we applied a similar methodology used in the determination of the distance-size ratio $D$: we proposed a set of exponential density functions (as the one in equation \ref{eq:density-exp}) with different values of $b$ and computed the difference between the gravity fields generated by the numerical model with the corresponding analytical solution, each time with a different value of $\delta$ ranging from 0.1 to 1 with a step of 0.1.
This tests have been performed for a thin and a thick spherical shell of 1km and 35km of thickness, respectively.
